Supplementary Information

### 2021-04-22 doc/manuscript/ForestGEO_treerings.Rmd
discussion, delete text on what method *could do*
#127
PZ comments

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
---

@@

Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent (Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].

~~Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, DBH, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].~~
~~This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, DBH and year, although other slowly-changing drivers could also be used.~~
We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @helcoski_growing_2019; @maxwell_declining_2016; @schurman_climatic_2019; @holz_radial_2018; @peltier_tree_2020; @zuidema_recent_2020].
In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, a conceptually parallel approach to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
For instance, the oldest trees, which are the only records available for the earliest decades, were the competitive winners, most likely with above-average growth rates [@aubry-kientz_joint_2015], thereby upwardly biasing growth rates in early decades [@groenendijk_no_2015].
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

@@

There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017]. 
With a few exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig. `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
A growth benefit of increasing atmospheric CO~2~ concentration is *potentially* expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.

>>>>

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
---

@@

Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig. `r clim_sens`). 
This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear. 
The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].

We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure such as ours that can account for this pattern and other DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.

@@

*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig. `r year_fig`).*
*These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, a conceptually parallel approach to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig. `r size`).
For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`, @brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
For instance, the oldest trees, which are the only records available for the earliest decades, were the competitive winners, most likely with above-average growth rates [@aubry-kientz_joint_2015], thereby upwardly biasing growth rates in early decades [@groenendijk_no_2015].
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

@@

There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017]. 
With a few exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig. `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
A growth benefit of increasing atmospheric CO~2~ concentration is *potentially* expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
